---
title: "HW 3"
author: "Mason Boyles"
date: "9/24/2024"
output: 
  html_document:
    number_sections: true
---

#

Let $E[X] = \mu$.  Show that $Var[X] := E[(X-E[X])^2] = E[X^2]-(E[X])^2$.  Note, all you have to do is show the second equality (the first is our definition from class). 

Distribute: E[(X-E[X])^2] = E(X^2 - 2XE[X] + (E[X])^2)

Expected Values: = E[X^2] - 2E[XE[X]] + E[(E[X])^2]

Considering E[X] as constant: = E[X^2] - 2E[X] * E[X] + (E[X])^2 * E[1]

Combining Terms: = E[X^2] - 2(E[X])^2 + (E[X])^2 

Combining terms: = E[X^2] - (E[X])^2

# 

In the computational section of this homework, we will discuss support vector machines and tree-based methods.  I will begin by simulating some data for you to use with SVM. 

```{r}
library(e1071)
set.seed(1) 
x=matrix(rnorm(200*2),ncol=2)
x[1:100,]=x[1:100,]+2
x[101:150,]=x[101:150,]-2
y=c(rep(1,150),rep(2,50))
dat=data.frame(x=x,y=as.factor(y))
plot(x, col=y)
```


##

Quite clearly, the above data is not linearly separable.  Create a training-testing partition with 100 random observations in the training partition.  Fit an svm on this training data using the radial kernel, and tuning parameters $\gamma=1$, cost $=1$.  Plot the svm on the training data.  

```{r}
set.seed(1)
train = sample(1:nrow(dat), 100)
train_data = dat[train,]

svmfit = svm(y~., data = train_data, kernel = "radial", gamma = 1, cost = 1, scale = FALSE)
plot(svmfit, train_data)

```

##

Notice that the above decision boundary is decidedly non-linear.  It seems to perform reasonably well, but there are indeed some misclassifications.  Let's see if increasing the cost ^[Remember this is a parameter that decides how smooth your decision boundary should be] helps our classification error rate.  Refit the svm with the radial kernel, $\gamma=1$, and a cost of 10000.  Plot this svm on the training data. 

```{r}

svmfit2 = svm(y~., data = train_data, kernel = "radial", gamma = 1, cost = 10000, scale = FALSE)
plot(svmfit2, train_data)

```

##

It would appear that we are better capturing the training data, but comment on the dangers (if any exist), of such a model. 

I would be afraid of overfitting in this case. It seems that the circle got split into two which may work to fit the training data well, but I fear it may not be the best at generalizing.

##

Create a confusion matrix by using this svm to predict on the current testing partition.  Comment on the confusion matrix.  Is there any disparity in our classification results?    

```{r}
#remove eval = FALSE in above
table(true=dat[-train,"y"], pred=predict(svmfit2, newdata=dat[-train,]))
```
It seems to have a relatively high number of situations where we predicted it to be a 2 but it was actually a 1. This is likely due to overfitting.

##

Is this disparity because of imbalance in the training/testing partition?  Find the proportion of class `2` in your training partition and see if it is broadly representative of the underlying 25\% of class 2 in the data as a whole.  

```{r}
sum(dat[-train,][,3] == 2) / nrow(dat[-train,])
```

No, .21 is broadly representative of the underlying 25% in the data as a whole.

##

Let's try and balance the above to solutions via cross-validation.  Using the `tune` function, pass in the training data, and a list of the following cost and $\gamma$ values: {0.1, 1, 10, 100, 1000} and {0.5, 1,2,3,4}.  Save the output of this function in a variable called `tune.out`.  

```{r}

set.seed(1)

tune.out = tune(svm, y~., data = train_data, kernal = "radial", ranges = list(cost = c(0.1, 1, 10, 100, 1000), gamma = c(0.5, 1, 2, 3, 4)))
tune.out


```

I will take `tune.out` and use the best model according to error rate to test on our data.  I will report a confusion matrix corresponding to the 100 predictions.  


```{r}
table(true=dat[-train,"y"], pred=predict(tune.out$best.model, newdata=dat[-train,]))
```

##

Comment on the confusion matrix.  How have we improved upon the model in question 2 and what qualifications are still necessary for this improved model.  

This improved performance significantly. We now have 2 less instances where we incorrectly classified a two as a one and 10 less instances where we incorrectly classified a one as a two. However, there still are significantly more instances where we are misclassifying ones as twos vs the other way around.

# 
Let's turn now to decision trees.  

```{r}

library(kmed)
data(heart)
library(tree)

```

## 

The response variable is currently a categorical variable with four levels.  Convert heart disease into binary categorical variable.  Then, ensure that it is properly stored as a factor. 

```{r}
heart$heart_disease = ifelse(heart$class == 0, 0, 1)
heart$heart_disease = as.factor(heart$heart_disease)
head(heart)

```

## 

Train a classification tree on a 240 observation training subset (using the seed I have set for you).  Plot the tree.  

```{r}
set.seed(101)

train=sample(1:nrow(heart), 240)

tree.heart = tree(heart_disease~. -class, heart, subset = train)
plot(tree.heart)
text(tree.heart, pretty = 0)

```


## 

Use the trained model to classify the remaining testing points.  Create a confusion matrix to evaluate performance.  Report the classification error rate.  

```{r}
tree.pred = predict(tree.heart, heart[-train,], type="class")
with(heart[-train,], table(tree.pred, heart_disease))

classification_error_rate = (3+8)/ (28+3+8+18)
classification_error_rate

```

##  

Above we have a fully grown (bushy) tree.  Now, cross validate it using the `cv.tree` command.  Specify cross validation to be done according to the misclassification rate.  Choose an ideal number of splits, and plot this tree.  Finally, use this pruned tree to test on the testing set.  Report a confusion matrix and the misclassification rate.  

```{r}
set.seed(101)
cv.heart = cv.tree(tree.heart , FUN = prune.misclass)
cv.heart
plot(cv.heart$size, cv.heart$dev, type = "b")
prune.heart = prune.misclass(tree.heart, best = 2) #Choose 2 because it is simpler than 4 even though both have the same dev value
plot(prune.heart)
text(prune.heart, pretty=0)

prune.pred = predict(prune.heart, heart[-train,], type="class")
with(heart[-train,], table(prune.pred, heart_disease))
classification_error_rate_pruned = ((9+8) / (9+8+28+12))
classification_error_rate_pruned
```


##

Discuss the trade-off in accuracy and interpretability in pruning the above tree. 

The pruned tree is far simpler, since there is only one decision that needs to be made. However, the performance is significantly worse in this training testing split as misclassification rate rose from ~.193 to ~.298

## 

Discuss the ways a decision tree could manifest algorithmic bias.  

One way that a decision tree can manifest algorithmic bias can be over-fitting caused by using a fully grown tree without pruning, this can cause it to not generalize well to data once it isn't exactly the same. Another way that is less specific to decision trees, but still applies is when you have unrepresentative training and testing data, they should have similar distributions of classes.